---
title: "TO 628 HW 1"
author: "Shanza Raza"
date: "10/22/2020"
output: 
  html_document:
    toc: true
    toc_float: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading my data 

```{r, cache=TRUE}
nyc <- read.csv("nycflights.csv")
class(nyc)

```

## Cleaning the data 

```{r}
# Deleting the column x
nyc$X <- NULL

# Factorize column that needs to be a factor 
nyc$origin <- as.factor(nyc$origin)
nyc$tailnum <- as.factor(nyc$tailnum)
nyc$month <- as.factor(nyc$month)
nyc$dest <- as.factor(nyc$dest)
nyc$carrier <- as.factor(nyc$carrier)
nyc$flight <- as.factor(nyc$flight)

#Recheck my data

str(nyc)

```

## Exploration of Data 

```{r}
#How many airlines are there?
uniquecarriers <- unique(nyc$carrier)
#print(uniquecarriers)
length(uniquecarriers)

#How many flights there were by the airline with code
codeOO <- nyc[nyc$carrier == "OO", ]
nrow(codeOO)

#How long is the shortest flight out of any NYC airport?
min(nyc$air_time, na.rm = TRUE)

#How many flights where there by United Airlines (code: UA) on Jan 12th 2013?
UAflights <- nyc[nyc$carrier == "UA" & nyc$year == 2013 & nyc$month == 01 & nyc$day == 12, ]
#print(UAflights)
nrow(UAflights)
```

## Arrival Delay

```{r}
#What was the average arrival delay for all airports and all airlines combined in Jan 2013?
AllDelayJan <- nyc[nyc$year == 2013 & nyc$month == 01, ]
#print(AllDelayJan)
mean(AllDelayJan$arr_delay, na.rm = TRUE) 

#Whats was the median arrival delay for all airports and all airlines combined in Jan 2013?
median(AllDelayJan$arr_delay, na.rm = TRUE) 

```


The distribution is skewed to the left based on the median value of `r mean(AllDelayJan$arr_delay, na.rm = TRUE)`  and mean value of `r median(AllDelayJan$arr_delay, na.rm = TRUE)` 

## Airline Performance
```{r}
#Calculate average arrival delays by airline?
tapply(nyc$arr_delay, nyc$carrier, mean, na.rm = TRUE)

#Draw a Bar Plot of Average Arrival Delays for all the Airlines 
AvgArrivalDelay <- tapply(nyc$arr_delay, nyc$carrier, mean, na.rm = TRUE)
barplot(AvgArrivalDelay)

#Which airline has the highest average arrival delay? Which airline has the smallest average arrival delay? Are there airlines that actually have negative average delay? 
min(AvgArrivalDelay)

names(AvgArrivalDelay[AvgArrivalDelay == min(AvgArrivalDelay)])

max(AvgArrivalDelay)

names(AvgArrivalDelay[AvgArrivalDelay == max(AvgArrivalDelay)])

```

The airline that has the highest average arrival delay is `r names(AvgArrivalDelay[AvgArrivalDelay == max(AvgArrivalDelay)])` and the airline that has the lowest average arrival delay is `r names(AvgArrivalDelay[AvgArrivalDelay == min(AvgArrivalDelay)])`. Yes, there are airlines that do have negative average delays such as `r names(AvgArrivalDelay[AvgArrivalDelay == min(AvgArrivalDelay)])` that has a delay of `r min(AvgArrivalDelay)` 

## Air Gain 

```{r}
nyc$AirGain <- nyc$dep_delay - nyc$arr_delay
#View(nyc)

#Explore airgain data and answer the question - do airlines actually gain any time when in air on average?
mean(nyc$AirGain, na.rm = TRUE)

#Calculate average airgain for different airlines
tapply(nyc$AirGain, nyc$carrier, mean, na.rm = TRUE)

CatchUp <- tapply(nyc$AirGain, nyc$carrier, mean, na.rm = TRUE)
barplot(CatchUp)
```


